---
title: "Supporting Information 2"
output:
  pdf_document: default
  html_document: default
date: "2025-09-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(dplyr)
require(magrittr)
require(gridExtra)
require(tidyr)
require(scales)
require(cowplot)
require(viridis)
require(broom)
require(formatR)
library(xfun)
library(GIFT)
library(purrr)
library(colorspace)
library(autoimage)
```

## Data preparation

First, prepare data inputs for BioenergeticHSC batch method 1, which implements the NREI model. This will basically create a large grid of depth, velocity, and temperature combinations. Other parameters (e.g., fish size) could also be modified here if desired.

```{r Data input, tidy.opts=list(width.cutoff=60)}
  velocity <- seq(1,100, by=2)
  depth <- seq(1, 100, by=2)
  temp <- seq(5, 23, by=1)
  hyd <- expand.grid(Velocity = velocity, Depth=depth)
  hyd_full <- bind_rows(replicate(19, hyd, simplify = FALSE))
  hyd_full$Temperature <- rep(temp, each = 2500) 
  hyd_full$Roughness <- ""
  hyd_full$ForkLength <- 6.5 ## Fork length in cm
  hyd_full$Mass <- 3 ## mass in g
  hyd_full$Turbidity <- 0
  hyd_full$DriftFile <-""
  hyd_full$Notes <- ""
  hyd_full$Label <- seq(1,47500, by=1)
  Batch1Input <- hyd_full %>% 
    dplyr::select(., Label, Depth, Velocity, Roughness, ForkLength, Mass, 
                  Temperature, Turbidity, DriftFile, Notes)
  
```

## Functions

The following functions will be used on the output of BioenergeticHSC to convert energetics to a daily timestep. First, the function *dGEI* will compute a daily gross energy intake (J), which factors in the costs of non-foraging activity (assumed to be equal to SMR). Foraging duration is assumed to be 14 h, but this could be adjusted. The bioenergetics parameters are for coho, but could be adjusted for other species.

```{r bioenergetics functions, tidy=TRUE}


#### Individual functions for energetics parameters
base_smr <- function(temp, W){ ## SMR (J/s) 
  SMR <- (1/3600) * exp(-2.94 + 0.87*log(W) + 0.064*temp)
  return(SMR*14.1*1.75)
}


Cmax <- function(temp, W){ ## Cmax (g/g/day) - Water temperature and weight (W)
  CTO <- 15  # Temperature corresponding to 0.98 of Cmax
  CTM <- 18  # Temperature at which temperature dependence is 0.98 of max
  CTL <- 26 # Upper temperature 
  CQ <- 5  # Lower water temperature 
  CK1 <- 0.42  # Unit-less fraction, see above
  CK4 <- 0.03  # Unit-less fraction, see above
  CA = 0.303  # intercept of the allometric function for a 1 g fish Topt
  CB = -0.275  # slope of allometric mass function
# Temperature dependence function
  G1 <-(1/(CTO-CQ)) * log((0.98*(1-CK1))/(CK1 * 0.02)) 	#1st
  L1 <- exp(G1*(temp - CQ))	#2nd 
  Ka <- (CK1 *L1) / (1 + CK1 * (L1-1))	#3rd 
  G2 <-(1/(CTL-CTM)) * log((0.98 * (1-CK4)) / (CK4 * 0.02))	#4th 
  L2 <-exp(G2*(CTL - temp)) 
  Kb <-(CK4*L2) / (1 + CK4* (L2-1)) 
  f_T <- Ka*Kb  
# Consumption equation. pCmax = 1, so drops out of equation 
  Cmax <- (CA * W^CB) * f_T
  return(Cmax)
}

SwimCost <- function(temp, W, velocity){ ## Swimming cost (J/s) 
  oq <- 14.1 ## Oxycaloric coefficient (Videler 1993)
  smr <- base_smr(temp, W)
  SC_chinook <- (1/3600)*oq*exp(-6.25 + (0.72 * log(W)) + (1.6 *log(velocity)))
  SC_coho <- (SC_chinook + (smr * 1.75)) ## Activity cost (J/s)
  return(SC_coho)
}

### Function that computes daily gross energy intake (J/day) for coho
### less resting costs while not foraging
dGEI <- function(GREI, temp, W){ 
  Cmax <- function(temp, W){ # Same Cmax function
    CTO <- 15  
    CTM <- 18  
    CTL <- 26 
    CQ <- 5  
    CK1 <- 0.42  
    CK4 <- 0.03  
    CA = 0.303  
    CB = -0.275  
    G1 <-(1/(CTO-CQ)) * log((0.98*(1-CK1))/(CK1 * 0.02)) 	
    L1 <- exp(G1*(temp - CQ))	
    Ka <- (CK1 *L1) / (1 + CK1 * (L1-1))	
    G2 <-(1/(CTL-CTM)) * log((0.98 * (1-CK4)) / (CK4 * 0.02))	
    L2 <-exp(G2*(CTL - temp)) 
    Kb <-(CK4*L2) / (1 + CK4* (L2-1)) 
    f_T <- Ka*Kb  
    Cmax_i <- (CA * W^CB) * f_T
    return(Cmax_i)
  }
## Energy costs O2/h while resting for coho (Trudel and Welch 2005)  
  SMR <- (exp(-2.94 + 0.87*log(W) + 0.064* temp))*1.75 
  GREI_total <- GREI * 14 * 3600 ## GREI (J/s) - Assume 14 h foraging window
## Cmax in total J. 3626 J/g wet weight from Waters (1977).   
  Cmax_J <- Cmax(temp, W) * 3626 * W 
## Total resting costs (J), assuming fish are resting in zero velocity habitats. 
## SMR converted to J from Videler 1993  
  Costs <- (SMR*14.1) * 10 
### Check if daily intake greater than Cmax, if it is then use Cmax
  DGEI <- ifelse(GREI_total < Cmax_J, GREI_total - Costs, Cmax_J - Costs)
  return(DGEI) 
}

### Same function but re-parameterized for rainbow trout (Railsback and Rose 1999; Rand 1993)
dGEI_rb <- function(GREI, temp, W){ 
  Cmax <- function(temp, W){ # Same Cmax function
    CTO <- 25  
    CTM <- 22.5  
    CTL <- 24.3 
    CQ <- 3.5 
    CK1 <- 0.2  
    CK4 <- 0.01  
    CA <- 0.303  
    CB <- -0.275  
    G1 <-(1/(CTO-CQ)) * log((0.98*(1-CK1))/(CK1 * 0.02)) 	
    L1 <- exp(G1*(temp - CQ))	
    Ka <- (CK1 *L1) / (1 + CK1 * (L1-1))	
    G2 <-(1/(CTL-CTM)) * log((0.98 * (1-CK4)) / (CK4 * 0.02))	
    L2 <-exp(G2*(CTL - temp)) 
    Kb <-(CK4*L2) / (1 + CK4* (L2-1)) 
    f_T <- Ka*Kb  
    Cmax_i <- (CA * W^CB) * f_T
    return(Cmax_i)
  }
## Energy costs O2/h for rainbow trout (Dodrill et al. 2016))  
  RA = 0.013
  RB = -0.217
  RQ = 2.2
  RTO = 22
  RTM = 26
  Y = log(RQ) * (RTM - RTO + 2)
  Z = log(RQ) * (RTM - RTO)
  X = (Z ** 2) * (1 + (1 + 40 / Y) ** 0.5) ** 2 / 400
  V = (RTM - temp) / (RTM - RTO)
  Rs_FT = V ^ X * exp(X * (1 - V))
  SMR <- (0.013 * (W ^ -0.217) * W * R_ft * 3420)/24 #resting costs per hour
  GREI_total <- GREI * 14 * 3600 ## GREI (J/s) - Assume 14 h foraging window
## Cmax in total J. 3626 J/g wet weight from Waters (1977).   
  Cmax_J <- Cmax(temp, W) * 3626 * W 
## Total resting costs (J), assuming fish are resting in zero velocity habitats. 
## SMR converted to J from Videler 1993  
  Costs <- (SMR*14.1) * 10 
### Check if daily intake greater than Cmax, if it is then use Cmax
  DGEI <- ifelse(GREI_total < Cmax_J, GREI_total - Costs, Cmax_J - Costs)
  return(DGEI) 
}

```

The input file is configured to predict habitat suitability on a 2D depth and velocity grid. However, for visualization, univariate curves may be preferred. The following functions extract depth or velocity curves where the maximum for each is realized (i.e., a 1D vector that intersects the optimum of the 2D suitability matrix). In other words, each depth or velocity curve assumes that the other variable is not limiting suitability. 
```{r }
## Function to extract HSCs
### Extracts rows that correspond to optimal depths (for velocity HSC)
HSC_Vextract <- function(x){  
  dd <- x$Depth[which.max(x$dHSI)]
  vhsc <- filter(x, Depth==dd)
  return(vhsc)
}

### Extracts rows that correspond to optimal depths (for depth HSC)
HSC_Dextract <- function(x){  
  vv <- x$Velocity[which.max(x$dHSI)]
  dhsc <- filter(x, Velocity==vv)
  return(dhsc)
}

```

## Data processing

```{r data processing}
##### Read in with batch method 1. Select and rename relevant columns
  bsc_input <- read.csv("./BioenergeticHSCbatch1output.csv")
  bsc <- na.omit(bsc_input)
  bsc_subset <- bsc %>% 
    dplyr::select(., Depth..cm., Velocity..cm.s., 
                  Fork.length..cm.,Mass..g., Temperature, 
                  Gross.rate.of.energy.intake..J.s.,
                        Total.energy.costs..J.s., )
  colnames(bsc_subset) <- c("Depth", "Velocity", "FL", "Mass", 
                            "Temperature", "GREI", "TotalCosts")

  
## Compute daily gross energy intake per day with dGEI function
  bsc_subset$dGEI <- 
    dGEI(bsc_subset$GREI, bsc_subset$Temperature, bsc_subset$Mass) 
## Compute total daily activity costs (swimming and maneuvering while foraging)
### Ensure hours match dGEI function. 14 h assumed here 
  bsc_subset$Costs_day <- bsc_subset$TotalCosts * 14 * 3600 
### Daily net energy intake  
  bsc_subset$dNREI <- bsc_subset$dGEI - bsc_subset$Costs_day 
  HSC_combined <- bsc_subset %>% 
    dplyr::select(., GREI, Depth, Velocity, FL, 
                                               Mass, Temperature, TotalCosts, 
                  dGEI, Costs_day, dNREI)  

## Proportion of Cmax  
  HSC_combined$Cmax_J <- 
    Cmax(HSC_combined$Temperature, HSC_combined$Mass) * 3626 * 3
  HSC_combined$pCmax <- HSC_combined$GREI*14*3600/HSC_combined$Cmax_J
  HSC_combined$pCmax <- if_else(HSC_combined$pCmax > 1, 1, HSC_combined$pCmax)
  
## Standardize to 0-1 scale. 
##All negative daily NREI is set to 0. 
##This data frame now contains temperature-dependent HSCs.
   HSC_combined$dHSI <- ifelse(HSC_combined$dNREI > 0, 
                               HSC_combined$dNREI/max(HSC_combined$dNREI), 0)  
   HSC_velocity <- HSC_Vextract(HSC_combined)
   HSC_depth <- HSC_Dextract(HSC_combined)

```

## Plotting
Code to reproduce figures from the manuscript
```{r fig1, warning=FALSE}
## Cmax across a range of temperature
  temp_range <- c(0:26)
  velocity <- c(0:50)
## Max consumption for 3 g fish across temperature range  
  C_max <- Cmax(temp_range, 3) 
## Make into data frame and convert to J  
  cdat <- data.frame(Temperature = temp_range, Cmax=C_max*3626) 
  cmax_plot <- cdat %>% ggplot(aes(x=Temperature, y=Cmax))+
    geom_line(lwd=1.25)+theme_bw()+ylab("Max Consumption (J/g/day)")+ xlab("Temperature (\u00b0C)")


## SMR across a range of temperatures    
  smr<- base_smr(temp_range, 3)
  smr_dat <- data.frame(SMR = smr, Temperature = temp_range)
  smr_plot <- smr_dat %>% ggplot(aes(x=Temperature, y=SMR))+
    geom_line(lwd=1.25)+theme_bw()+ylab("Standard Metabolic Rate (J/s)")+
    xlab("Temperature (\u00b0C)")
  
## Activity costs across velocity and temperatures
  temp_range_small <- c(6,12,18,24) 
  tempxvelocity <- expand.grid(velocity, temp_range_small)
  names(tempxvelocity) <- c("Velocity", "Temperature")
  tempxvelocity$cost <- 
    SwimCost(tempxvelocity$Temperature, 3, tempxvelocity$Velocity)
  tempxvelocity$Temperature <- factor(tempxvelocity$Temperature)
  activity_plot <- 
    tempxvelocity %>% ggplot(aes(x=Velocity, y=cost, colour=Temperature))+
    geom_line(lwd=1.25)+theme_bw()+
    ylab("Activity Costs (J/s)") + xlab("Velocity (cm/s)") +
    scale_colour_manual(values=c("blue", "yellow","orange", "red"))
  
## Combine into one figure 
fig1 <- grid.arrange(smr_plot, activity_plot,cmax_plot, nrow = 2, 
                layout_matrix= rbind(c(1,2), 3))
ggsave("./Figures/Fig1.pdf", fig1,
           width = 10, height = 8, units = "in", dpi = 600)
  
```

### Figure 2

```{r fig2, warning=FALSE}
## Illustrate modelling process and the influence of Cmax on daily NREI


  HSC_velocity$Cmax_12 <- Cmax(12, HSC_velocity$Mass)*3626*3
  HSC_velocity$Cmax_6 <- Cmax(6, HSC_velocity$Mass)*3626*3
  
## Plots of energy budgets against velocity at different temperatures
  
  example_6C <- HSC_velocity %>% filter(., Temperature == 6) %>% ggplot()+
    geom_line(aes(x=Velocity, y=dGEI), color="green", lwd=1.2)+
    geom_line(aes(x=Velocity, y=Costs_day), color="red", lwd=1.2) + 
    geom_line(aes(x=Velocity, y=dNREI), color="black", lwd=1.2) +
    geom_line(aes(x=Velocity, y=Cmax_6), lty=2, lwd=1.2)+
    theme_bw() + ylim(0,2500)+ xlim(0,50)+ylab("Energy (J/day)") + 
    xlab("Velocity (cm/s)")

  
  example_12C <- HSC_velocity %>% filter(., Temperature == 12) %>% ggplot()+
    geom_line(aes(x=Velocity, y=dGEI), color="green", lwd=1.2)+
    geom_line(aes(x=Velocity, y=Costs_day), color="red", lwd=1.2) + 
    geom_line(aes(x=Velocity, y=dNREI), color="black", lwd=1.2) +
    geom_line(aes(x=Velocity, y=Cmax_12), lty=2, lwd=1.2)+
    theme_bw() + ylim(0,2500)+ xlim(0,50)+ylab("") + 
    xlab("Velocity (cm/s)") + theme(axis.text.y = element_blank())
  
fig2 <-  plot_grid(example_6C, example_12C, 
                  labels = c("6\u00b0 C", "12\u00b0 C"), label_x=0.5, label_y = 0.99)
ggsave("./Figures/Fig2.pdf", fig2,
          width = 10, height = 8, units = "in", dpi = 600)

```

### Figure 3
```{r fig3, warning=FALSE}
### Prepare data frames with univariate curves across temperatures
  HSC_velocity$Temperature <- factor(HSC_velocity$Temperature)
  HSC_depth$Temperature <- factor(HSC_depth$Temperature)
  
### Velocity plot    
  vhsi_plot <- HSC_velocity %>% 
    ggplot(aes(x=Velocity, y=dHSI, colour=Temperature))+
    geom_line(se=F, lwd=0.75)+theme_bw()+ylim(0,1)+
    xlim(0,40)+xlab("Velocity (cm/s)") + 
    ylab("Suitability (Standardized J/day)")+ 
    scale_colour_viridis(discrete = TRUE)+theme(legend.position = "none")

### Depth plot  
  dhsi_plot <- HSC_depth %>% ggplot(aes(x=Depth, y=dHSI, colour=Temperature))+
    geom_smooth(se=F, lwd=0.75)+
    theme_bw()+ylim(0,1)+xlim(0,100)+xlab("Depth (cm)") + 
    ylab(NULL)+ scale_colour_viridis(discrete = TRUE)+
    theme(axis.text.y=element_blank(),axis.ticks.y = element_blank())
  



fig3 <- plot_grid(vhsi_plot, dhsi_plot + 
guides(colour = guide_legend(override.aes = list(linewidth = 1.5))))
#   ggsave("./Figures/Fig3.pdf", fig3,
#           width = 10, height = 8, units = "in", dpi = 600)
  
```

### Figure 4
```{r fig4, warning=FALSE}
### Proportion of Cmax at the optimal depth/velocity across temperature
  pcmax <- HSC_velocity %>% group_by(Temperature) %>%
    filter(dHSI == max(dHSI)) %>% arrange(Temperature) %>%
    mutate(Temperature = as.numeric(Temperature))%>%
    mutate(ration = Cmax_J * pCmax)
    
   
   pcmax_plot <- pcmax %>% ggplot()+
    geom_hline(yintercept = 1, color="red", lty=2, lwd=2)+
    geom_point(aes((Temperature+4), pCmax,), size=3)+theme_bw()+
     ylab("Proportion of Cmax")+xlab("") + ylim(0.5,1)+
     theme(axis.text.x=element_blank(),axis.ticks.x = element_blank())
  
  

   ration_plot <- pcmax %>% ggplot()+
     geom_line(aes((Temperature+4), Cmax_J), colour="red", lwd=2) +
     geom_point(aes((Temperature+4), ration), size=3)+theme_bw()+
    ylab("Daily Ration (J)") + xlab("Temperature (\u00b0C)")

 fig4 <- plot_grid(pcmax_plot, ration_plot, nrow = 2, align = "v", labels = c("a", "b"))
 ggsave("./Figures/Fig4.pdf", fig4,
           width = 10, height = 8, units = "in", dpi = 600)

```


## Simple example application
```{r Worked examples}

### Simulate channel hydraulics across flows with hydraulic geometry simulator
  hyd <- AvgHydraulics(S=0.01, wb=10, db=0.5, db_max=NULL, D84=100, xs_output=TRUE)


### Generate flow x
  qtemp <- data.frame(Q = c(0.10, 0.08, 0.05, 0.05, 0.05), temperature = c(20, 20, 21, 22, 22), Time=c(1,2,3,4,5), Scenario = "Base")
  
  qtemp_fa <- data.frame(Q = c(0.10, 0.09, 0.09, 0.09, 0.09), temperature = c(20, 20, 21, 22, 22), Time=c(1,2,3,4,5), Scenario = "Flow intervention")
  
  qtemp_ta <- data.frame(Q = c(0.10, 0.09, 0.09, 0.09, 0.09), temperature = c(20, 19, 18, 17, 17), Time=c(1,2,3,4,5), Scenario = "Flow and temperature intervention")
  
  qt_scenarios <- rbind.data.frame(qtemp, qtemp_fa, qtemp_ta)

### Generate WUA predictions across all temperatures  
  HSC_velocity <- HSC_velocity %>% mutate(., Velocity = Velocity*0.01) # convert to m/s
  HSC_depth <- HSC_depth %>% mutate(., Depth = Depth*0.01) # convert to m

## Group into depth and velocity HSI
  velocity_list <- HSC_velocity %>% group_by(Temperature)%>% select(., Velocity, dHSI) %>% rename(., velocity=Velocity, suit=dHSI) %>% group_split()
  depth_list <- HSC_depth %>% group_by(Temperature)%>% select(., Depth, dHSI) %>% rename(., depth=Depth, suit=dHSI) %>% group_split()

## Apply Habitat function to generate WUA prediction for each temperature
  
habsim <- function(d, v){
  Habitat(hyd, d_curve = d, v_curve = v) ## Habitat function from GIFT package
}  

wua_list <- Map(habsim, depth_list, velocity_list)
wua_df <- list_rbind(wua_list, names_to = "temperature")
wua_df <- wua_df %>% mutate(., temperature = temperature+4) ## offset to correct temperatures
wua_df$temperature_level <- factor(wua_df$temperature)

## Irregular Q intervals in hydraulic output 

wua_smooth <- function(df){
model <- loess(WUA ~ Q, data=df, span=0.3)
smooth_data <- augment(model, df)
return(smooth_data)
}

wua_df <- wua_df %>% group_by(temperature_level) %>% group_map(~wua_smooth(.), keep = TRUE) %>% bind_rows(.)

### Match predicted WUA with flow and temperature time series
qtemp_wua <- left_join(qt_scenarios, wua_df, by = c("Q", "temperature"))



```

### Figure 5
```{r}



wua_heatplot <- wua_df %>% filter(., Q < 0.10) %>%  ggplot(., aes(x = Q, y = temperature)) + geom_raster(aes(fill=.fitted), interpolate = TRUE) + 
scale_fill_gradient(low = "lightblue", high = "red")+xlab("Flow (CMS)") + ylab("Temperature (\u00b0C)")+theme_bw()+labs(fill = "WUA")  

tplot <- qtemp_wua %>% ggplot(., aes(x=Time, y = temperature, color=Scenario))+
  geom_line(linewidth=1.5)+theme_bw()+ ylim(10,24)+ylab("Temperature (\u00b0C)") + theme(axis.text.x=element_blank())+theme(legend.position = "none")+
   theme(plot.margin = unit(c(0.1, 0.1, 0, 0.5), "cm")) + xlab("")

qplot <- qtemp_wua %>% ggplot(., aes(x=Time, y = Q, color=Scenario))+
  geom_line(linewidth=1.5)+theme_bw() + ylab("Flow (CMS)")+ theme(axis.text.x=element_blank())+theme(legend.position = "none") +
  theme(plot.margin = unit(c(0.1, 0.1, 0, 0.5), "cm")) + xlab("")
  
  wuaplot <- qtemp_wua %>% ggplot(., aes(x=Time, y = WUA, color=Scenario))+
  geom_line(linewidth=1.5)+theme_bw()+theme(legend.position = "bottom")+
    theme(plot.margin = unit(c(0.1, 0.1, 0.1, 0.5), "cm")) + theme(legend.title = element_blank()) 


  fig5 <- plot_grid(wua_heatplot, tplot, qplot, wuaplot, nrow=4, ncol=1, rel_heights = c(2,1,1,1.5), align = "v", axis = "l", labels=c("a", "b", "c", "d"))
  
ggsave("./Figures/Fig5.pdf", fig5,
         width = 8, height = 10, units = "in", dpi = 600)

```



```{r S2 figures}
ui <- hyd %>% ggplot(., aes(x=Q, y=Ui)) +
  geom_line(linewidth=1.5) + theme_bw() + xlab("") + ylab("Velocity (m/s)")+
  xlim(0,1.5) + ylim(0,1)

di <- hyd %>% ggplot(., aes(x=Q, y=di)) +
  geom_line(linewidth=1.5) + theme_bw() + xlab("") + ylab("Depth (m)")+
  xlim(0,1.5) + ylim(0,0.5)


wi <- hyd %>% ggplot(., aes(x=Q, y=Wi)) +
  geom_line(linewidth=1.5) + theme_bw() + xlab("Flow (CMS)") + ylab("Wetted Width (m)") + xlim(0,1.5)
  

figS1 <- plot_grid(ui, di, wi, nrow=3, ncol=1)

ggsave("./Figures/FigS1.tiff", figS1,
         width = 10, height = 8, units = "in", dpi = 600)

```